cardinalR: Generating interesting high-dimensional data structures

### 2025-05-06 paper-cardinalR.Rmd
Added branching definitions

@@

>>>>

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in a $p$-dimensional space. Each branch lies primarily within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible $2\text{-}D$ combinations. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(x\_{i1}, x\_{i2})$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n\_i$ points are generated, where $\sum\_{i=1}^k n\_i = n$. The structure follows:

$$
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5),
$$

where $s\_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s\_i = 1$. When sampling with replacement, $s\_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$.

For $p > 2$, the remaining dimensions contain independent Gaussian noise:

$$
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$

@@

The `gen_orgcurvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in a $p$-dimensional space. Each branch is constructed in a unique or repeated $2\text{-}D$ subspace of the $p$-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining $p - 2$ dimensions contain Gaussian noise.

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(x\_{i1}, x\_{i2})$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s\_i$ is used to vary the curvature:

* If $k \leq \binom{p}{2}$: $s\_i = 1 $
* If $k > \binom{p}{2}$: $s\_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

Each branch contains $n\_i$ points such that $\sum\_{i=1}^k n\_i = n$, where the vector $(n\_1, \dots, n\_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$.

Within its assigned subspace $(x\_{i1}, x\_{i2})$, branch $i$ is defined by:

$$
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5)
$$

This forms a smooth downward-opening parabola in the plane defined by $(x\_{i1}, x\_{i2})$, with the degree of curvature controlled by $s\_i$.

For $p > 2$, all remaining dimensions are independent Gaussian noise:

$$
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$
